Decoding Liver Cancer Prognosis: From Multi-omics Subtypes, Prognostic Models to Single Cell Validation

Version 1

Abstract

Abstract

Purpose

Hepatocellular carcinoma (HCC) is a highly aggressive tumor characterized by significant heterogeneity and invasiveness, leading to a lack of precise individualized treatment strategies and poor patient outcomes. This necessitates the urgent development of accurate patient stratification methods and targeted therapies based on distinct tumor characteristics.

Experimental Design

By integrating gene expression data from The Cancer Genome Atlas (TCGA), International Cancer Genome Consortium (ICGC), and Gene Expression Omnibus (GEO), we identified subtypes through a multi-omics consensus clustering approach amalgamated from 10 clustering techniques. Subsequently, we developed a prognostic model, employing machine learning algorithms, based on subtype classification features. Finally, by analyzing single cell sequencing data, we investigated the mechanisms driving prognostic variations among distinct subtypes.

Results

First, we developed a novel consensus clustering method that categorizes liver cancer patients into two subtypes, CS1 and CS2. Second, we constructed a prognostic prediction model, which demonstrated superior predictive accuracy compared to several models published in the past five years. Finally, we observed differences between CS1 and CS2 in various metabolic pathways, biological processes, and signaling pathways, such as fatty acid metabolism, hypoxia levels, and the PI3K-AKT signaling pathway.

Introduction

Liver cancer is currently one of the most common tumors, and among primary liver cancers, hepatocellular carcinoma (HCC) is the most common pathological type. In 2022, liver cancer saw widespread new cases globally, and its incidence rate has been gradually increasing in recent years, ranking 6th among all tumors worldwide[1]. Meanwhile, liver cancer ranks 3rd among tumors causing patient deaths, with a relative 5-year survival rate of about 18%, highlighting the significantly poor prognosis of this disease[2].

With the advancement of current treatment technologies, there are ongoing attempts to employ advanced treatment methods for innovative liver cancer therapy. Nevertheless, up to now, the progress of strategies aimed at treating liver cancer, especially unresectable liver cancer, has been limited. For instance, targeted therapies and immunotherapies have been the focus of extensive research efforts and have shown promising results in some patients. However, a significant number of patients still do not obtain clinical benefits from these approaches. This may be due to the significant heterogeneity of liver cancer; classifying liver cancer patients according to molecular subtypes might help overcome this difficulty[3]. This urgently necessitates leveraging large-scale multi-omics data and advanced machine learning algorithms to identify biomarkers capable of efficiently predicting clinical treatment outcomes for liver cancer patients.

In this study, we combined mRNA, long non-coding RNA (lncRNA), and microRNA (miRNA) expression profiles, genomic mutations, and epigenomic DNA methylation data, using a consensus strategy integrating 10 multi-omics clustering methods to develop comprehensive consensus subtypes of liver cancer, and conducted preliminary exploration of the molecular characteristics and potential clinical applications of the two subtypes. Subsequently, we further expanded the clinical application value of the subtype. Based on the differential gene expression profiles of different subtypes, we used a combination of 101 machine learning algorithms to construct the best-performing prognostic model. This prognostic model showed excellent prognostic value in both training and validation cohorts, and outperformed several other prognostic models in comparisons. Finally, we validated the subtypes at the single-cell sequencing cell level using the same typing method, and further explored the possible causes and mechanisms of prognostic differences in patients with different subtypes.

Materials and methods

Data collection and processing

Download the molecular profiles of the TCGA-LIHC cohort dataset from the TCGA database, including 432 LIHC cases with complete transcriptome expression, somatic mutations, copy number alterations (CNA), DNA methylation, and clinical information. The transcriptome profiles of mRNA and lncRNA were obtained using the “TCGAbiolinks” R package[4], and GENCODE27 was used to convert Ensemble IDs to gene symbols; Somatic mutation data were also obtained using “TCGAbiolinks” R package and processed with the “maftools” R package[5]; DNA methylation profiles and clinical information were downloaded from UCSC Xena (https://xenabrowser.net/); Meanwhile, we also downloaded 445 LIHC cases with complete transcriptome expression and clinical information from the LIHC.JP cohort in the ICGC database (https://dcc.icgc.org/); The single-cell sequencing dataset is sourced from the GEO database (https://www.ncbi.nlm.nih.gov/geo/, accession number GSE229772).

Construct multi-omics consensus clustering classification and perform validation

In this study, we utilized the “getElites” function from the multi-omics integration and visualization (MOVICS) package to screen gene features in cancer subtypes[6]. For continuous variables (mRNA, lncRNA, miRNA, and methylation), we set the “method” parameter of the “getElites” function to “mad” to screen the top 1000 genes with the greatest variability; For binary variable gene mutation data, we conducted screening based on mutation frequency by setting the “method” parameter to “freq” to identify genes with a mutation frequency exceeding 10. The results from these five dimensions were incorporated into our study for further analysis. After the initial feature selection, we used the “getClustNum” function in the MOVICS package to further determine the optimal number of clusters for our study and ultimately decided to divide them into two subtypes. Next, we applied the “getMOIC” function for clustering analysis. We used 10 clustering algorithms (CIMLR, ConsensusClustering, SNF, iClusterBayes, PINSPlus, moCluster, NEMO, IntNMF, COCA, and LRA) as input for the “methods.list” parameter and employed the default parameters provided by the MOVICS package. As a result, we obtained the clustering results for each method. After calculating the clustering results of the 10 methods, we integrated the results of different algorithms based on the concept of consensus clustering using the “getConsensusMOIC” function, obtaining the final clustering results. Subsequently, we conducted nearest template prediction(NTP) analysis, performed subtyping in the external ICGC-JP cohort, and validated the subtyping method.

Molecular Characteristics of Multi-Omics Consensus Clustering Subtypes

We analyzed the mutation status using the “maftools” R package, then used the limma method to analyze the differentially expressed genes between the two subtypes[7], and performed enrichment analysis with the clusterProfiler R package[8].

Cell, signaling pathways, stemness, and other phenotypic characteristics of multi-omics consensus clustering subtypes

Using ssGSEA analysis, we calculated the enrichment scores of different immune cell characteristics between the two subtypes based on immune cell gene sets reported in the literature[9]; Next, we analyzed the activation status of cancer pathways and some biological processes in patients using the GSEA algorithm, and analyzed the sensitivity of the two subtypes to conventional therapeutic drugs.

Both of these steps were performed using the “GSVA” R package[10]. The expression data from CCLE were sourced from the Broad Institute Cancer Cell Line Encyclopedia (https://sites.broadinstitute.org/ccle/); The CTRP v.2.0 (https://portals.broadinstitute.org/ctrp) and PRISM dataset (24Q2; https://depmap.org/portal/prism/) were utilized to acquire drug sensitivity data; The half-maximal inhibitory concentration (IC50) was used as an indicator of drug sensitivity.

Constructing a prognostic prediction model based on the consensus clustering of multi-omics features

To build the prognostic model, we randomly split the TCGA-LIHC cohort into training and validation sets at a 1:1 ratio and incorporated 10 machine learning algorithms, including CoxBoost, stepwise Cox, Lasso, Ridge, Enet, survival-SVMs, GBMs, SuperPC, plsRcox, and RSF, creating a total of 101 different combinations of these algorithms[11]. Ultimately, these 101 algorithm combinations will be utilized to develop the most predictive prognostic model with the best C-index performance. We then scored each sample in the training and validation sets based on the model and classified the samples into high-risk and low-risk groups according to the optimal cut-off value of scores, and the “survminer” R package(https://github.com/kassambara/survminer/tree/master) was used to determine the optimal cut-off value. Cox regression and Kaplan–Meier analyses were performed via the “survival” R package(https://CRAN.R-project.org/package=survival).

Comparison of Prognostic Prediction Models

We evaluated the prognostic predictive ability of all models included in the study using the C-index in each cohort. To enhance the clinical utility of the model, we constructed a nomogram based on factors obtained from multivariable Cox regression. We plotted time-dependent C-index curves and calibration curves to describe accuracy and used decision curve analysis to calculate the clinical benefit for patients.

scRNA-seq Data Processing

We utilized the “Seurat” R package to process single-cell sequencing data from the GSE229772 dataset[12]. We removed samples with over 10% mitochondrial genes and over 5% ribosomal genes, and filtered out genes expressed in fewer than 3 cells. Next, we identified the most variable genes using the FindVariableFeatures function from the Seurat R package, conducted dimensionality reduction via principal component analysis (PCA), and identified a total of 13 clusters through t-SNE clustering analysis. Finally, we annotated the cell populations by combining the expression of characteristic genes in each cell cluster with the expression of commonly used cell-type-specific genes across the clusters.

Characteristic Analysis of Different Cell Subpopulations

We used the “irGSEA” R package[13], employing 6 methods (AUCell, UCell, singscore, ssGSEA, JASMINE, and viper) for pathway enrichment analysis among different cell populations, and performed a comprehensive evaluation of the differential analysis results using robust rank aggregation (RRA) to identify gene sets significantly enriched in most enrichment analysis methods. Moreover, we additionally utilized the “scMetabolism” R package to compare the metabolic difference across various cell types[14]. The metabolic activity of each cell was quantified through integrating the gene sets contained in the KEGG database. Subsequently, we used the “CellChat” R package to explore the interaction patterns among cell populations[15], defining ligands and receptors as outgoing and incoming signals, respectively.

Characteristics of Different Subtypes of Malignant Cells at the Single-Cell Level

We applied the “CopyKAT” R package to analyze gene expression levels across all cells[16], identifying diploid cells as normal cells, with normal and tumor cells separated due to differences in gene expression distributions. Subsequently, we classified tumor cells based on the previously obtained multi-omics consensus clustering approach, and used the irGSEA R package to perform pathway enrichment analysis on different subtypes, conducted metabolic differential analysis with the scMetabolism R package, and ultimately explored the interaction patterns of different subtype malignant cells with other cell populations using the CellChat package.

Results

Construct multi-omics consensus clustering classification and perform validation

We utilized clustering prediction indicators, gap statistical analysis, silhouette scores, and insights from previous research involving 10 multi-omics integrated clustering algorithms to ascertain the number of subtypes (Fig.1A). Using a consensus integration method, we linked the clustering results to distinct molecular expression patterns of the transcriptome (mRNA, lncRNA, and miRNA), epigenetic methylation, and somatic mutations, thereby identifying two subtypes (CS1 and CS2) independently (Fig.1B). Significant differences in mRNA, lncRNA, miRNA, DNA CpG methylation sites, and mutation gene characteristics were observed between the two subtypes (Fig.1C). This classification method is strongly correlated with overall survival (p < 0.001), with patients of the CS2 subtype exhibiting better prognostic survival compared to those of the CS1 subtype (Fig.1D). These characteristics suggest that our classification method can effectively distinguish between patients. Subsequently, employing the NTP method, we selected 200 genes specifically upregulated in each subtype as classifiers to validate the classification in an independent external validation set, ICGC, and found a high degree of consistency in the classification results. Patients with CS1 and CS2 subtypes in the ICGC cohort exhibited similar prognostic differences (Fig.1E-G). This demonstrates the versatility and practical value of our classification method.

Multi-omics consensus clustering classification. (A) Calculation of optimal number of subtypes based on cluster prediction index and gap-statistics. (B) Heatmap of the consistency of the integration of the results of the 10 typing algorithms. (C-D) Significant multiomics characterization and prognostic differences between subtypes. (E-G) Validation of consistency of typing results and prognostic value in an external cohort.

Molecular Characteristics of Multi-Omics Consensus Clustering Subtypes

We analyzed the gene mutation landscape of two subtypes and found that both subtypes exhibited a high frequency of mutations in TP53, KMT2D, and DNAH6 genes(Fig.2A). Based on the correlation between tumor mutation burden (TMB) and response to immunotherapy, we initially evaluated the TMB of patients within the two subtypes. The results indicated no significant difference in TMB between the two subtypes (Fig.2B-C). Subsequently, we analyzed the bulk RNA sequencing data of patients within the two subtypes, identified differentially expressed genes, and conducted enrichment analysis on these genes. The results revealed significant differential enrichment in several biological pathways, including those related to metabolism and signal transduction, such as lipid peroxidation, fatty acid metabolism, and amino acid metabolism, suggesting notable biological differences between the tumors and associated tissues of the two subtypes (Fig.2D-E). Previous studies have reported that heterogeneity among different tumor tissues of the same tumor type largely stems from metabolic reprogramming and changes in signal transduction pathway activity[17–18]. This implies that the heterogeneity between the tumor tissues of the two subtypes may also arise from these factors. Finally, we compared the consistency of subtyping methods with the pathologic staging of patients(Fig.2F).

Differences in molecular characterization of subtypes. (A) Differences in mutation landscapes of subtypes. (B-C) Genomic characterization from TMB and FGA of subtypes. (D-E) Functional enrichment of up and down-regulated differential genes between subtypes. (F) Concordance between subtypes and patients’ pathologic staging.

Cell, signaling pathways, stemness, and other phenotypic characteristics of multi-omics consensus clustering subtypes

Given the impact of the tumor immune microenvironment on tumor tissue heterogeneity and patient treatment responsiveness, we used the ssGSEA algorithm to assess the enrichment status of immune-related cells in the tumor microenvironment (TME) of two subtypes(Fig.3A). The gene set utilized for this analysis was compiled and summarized from various literature sources. Then, we employed drug intervention data to evaluate the sensitivity of the two subtypes to commonly used cancer therapies. The CS2 subtype exhibited higher sensitivity to 5-fluorouracil, Gefitinib, Afatinib, and Erlotinib compared to the CS1 subtype (Fig.3B). Among these, 5-fluorouracil is the main drug used in hepatic arterial infusion chemotherapy for hepatocellular carcinoma (HCC) patients, whereas Gefitinib, Afatinib, and Erlotinib are tyrosine kinase inhibitors (TKIs) that primarily inhibit epidermal growth factor receptor (EGFR). We further compared the level of infiltration of each cell taxon in subtypes in the TME(Fig.3C). Within the TME of the CS2 subtype, associated with a comparatively more favorable prognosis, nearly all assessed cell distribution levels demonstrated a decrease in comparison to those in the CS1 subtype. This encompasses traditionally recognized inhibitory immune cells like MDSCs and Regulatory T cells, alongside pivotal tumor-killing immune cells like Activated CD8+ T cells and NK cells. These findings may imply reduced levels of tumor-associated antigens within CS2 tumor tissue, leading to a comparatively subdued activation and induction of immune cells within the TME. We also evaluated several common biological pathways that may influence the TME and tumor tissue responsiveness to treatment. We discovered that the CS1 subtype is more active than the CS2 subtype in multiple pathways associated with tumor development and progression, including the MAPK, JAK-STAT, NFκB, PI3K-AKT, and VEGF signaling pathways, as well as pathways involved in angiogenesis and hypoxia. No significant differences were observed in the cAMP signaling pathway or the mRNAsi stemness score (Fig. 3D).

Phenotypic features of subtypes that may be associated with tumor development. (A) Landscape of subtype TME immune infiltration. (B) Differences in subtype responsiveness to different drugs. (C) Comparison of immune cell infiltration levels in subtypes of TME. (D) Comparison of hepatocellular carcinoma-related signaling pathways and biological process activity in subtypes

Constructing a prognostic prediction model based on the consensus clustering of multi-omics features

In a previous study, we utilized the mRNA expression profiles of TCGA-LIHC from the TCGA database to compute the differential gene expression profiles between CS1 and CS2 subtypes. We selected the top 30 significantly different genes as candidate genes for prognostic risk. Subsequently, we applied 101 algorithmic approaches to determine the correlation and impact weights between these 30 genes and prognosis, leading to the construction of a prognostic prediction model. Simultaneously, we calculated the average C index of each model to assess the predictive capability across all models (Fig. 4A). Employing this model to evaluate patients’ prognostic risk within the training set TCGA-LIHC cohort and stratifying them based on the optimal cutoff value, we observed a notably poorer prognosis among high-risk group patients compared to their low-risk counterparts. The model comprises 5 central genes, and the Hazard Ratio associated with these genes in the TCGA-LIHC cohort is depicted in Fig. 4C. Additionally, we illustrated the patients’ prognostic outcomes and variations in central genes as the prognostic risk escalated (Fig. 4D). Subsequently, we validated the model using an independent external validation set, the ICGC cohort. Our findings demonstrated its efficacy in stratifying patients into high and low-risk groups within the ICGC cohort. A notable disparity in prognosis between the two groups was evident, aligning with the outcomes observed in the training set (Fig. 4E-F).

Building prognostic prediction models through machine learning. (A) 101 Machine Learning Algorithms to Build Screening Optimal Prognostic Models. (B) Calculating risk scores and comparing prognostic differences based on predictive modeling. (C-D) Correlation between 5 prognostic genes and prognosis. (E-F) Validating prognostic model efficacy in an external cohort external cohort.

Comparison of Prognostic Prediction Models

With the rapid advancement and widespread adoption of Next Generation Sequencing (NGS) technology, numerous prognostic prediction models based on gene expression have been extensively reported. After developing our model, we reviewed and collected prognostic prediction models reported in the past five years and included seven of these models, which are based on various foundations such as immune infiltration, glucose metabolism, and lipid metabolism. We calculated the C-index for a total of eight models using the training set TCGA-LIHC cohort and the validation set ICGC cohort. The results indicated that our model demonstrated superior prognostic prediction performance compared to all other models in both cohorts (Fig. 5A). Subsequently, we incorporated clinical features related to prognosis into the model to construct a nomogram (Fig. 5B). The calibration curve demonstrated that the nomogram had predictive accuracy more consistent with actual conditions (Fig. 5C). Decision curve analysis (DCA) indicated that both the nomogram and the original model could benefit patients, with the nomogram performing slightly better (Fig. 5D).

Comparison and refinement of prognostic prediction models. (A) Comparison of predictive performance of prognostic models with some prognostic models published in the last 5 years. (B) Constructing a nomogram incorporating patient clinical characteristics. (C-D) Evaluating model performance based on calibration curve and decision curve analysis.

scRNA-seq Data Processing

Given the substantial prognostic variances between CS1 and CS2 patients and the notable heterogeneity in tumor tissues at both cellular and molecular levels, we sought to explore the potential mechanisms driving these differences through scRNA-seq data analysis. The GSE229772 dataset from the GEO database was chosen for our analysis. Following quality control filtering, 56,938 cells were derived from 31 samples. By employing hierarchical and unsupervised clustering methods, we delineated cell subclusters according to their expression profiles, categorizing the cells into 13 subclusters (Fig. 6A). Subsequently, we employed established marker genes to profile various cell types and assign annotations, discerning 12 primary cell subgroups: B cells, stromal cells, vascular endothelial cells, CD8+ T cells, epithelial cells, fibroblasts, germ cell-like cells, hepatic stellate cells, hepatocytes, macrophages, neutrophils, and T cells (Fig. 6B-D). Within these, the immune cell group prevailed as the most abundant, succeeded by stromal cells and fibroblasts (Fig. 6E).

ScRNA-seq cell sorting and annotation. (A-C) Cellular subpopulation delineation and annotation based on cellular characterization genes. (D-E) Visualization of the proportions of each cell subgroup. (F) Differentially expressed genes by cellular subpopulations

Characteristic Analysis of Different Cell Subpopulations

To investigate pathway heterogeneity among different cell subclusters, we identified differentially expressed characteristic genes within these subclusters. Subsequently, we employed multiple algorithms, including AUCell, UCell, singscore, ssgsea, JASMINE, and viper, to assess pathway activity variation. Differential analysis results were comprehensively evaluated using robust rank aggregation (RRA), and gene sets significantly enriched across multiple enrichment analysis methods were filtered to obtain consensus results. Statistically significant differentially enriched pathways from these consensus results were visualized (Fig.7A-B). We observed significant differences in several metabolic and signaling pathways among different cell subclusters. Given the reported impact of metabolic changes and signaling pathway variations on the TME and their implications for heterogeneity, we specifically analyzed these pathways. Utilizing the VISION algorithm, we assessed metabolic levels across each cell subcluster and visualized the differences in a dot plot (Fig.7C). Our findings indicated pronounced metabolic variations among cell subclusters, with germ cell-like cells and hepatocytes notably exhibiting heightened metabolic activity compared to other cell groups. Subsequently, we utilized the CellChat R package’s integrated databases for secreted signaling, ECM receptor signaling, cell-cell contact, and non-protein signaling to explore cell interactions within the single-cell dataset. Initially, we examined the cell communication landscape between distinct cell groups and constructed a cell communication network (Fig.7D). Our analysis revealed robust communication links between various cell groups, particularly notable between fibroblasts and hepatic stellate cells, which exhibited signaling connections with nearly all other cells. Then, we visualized the signaling output and input patterns (Fig.7E-G) to illustrate the contributions of different cell groups at both ends.

Phenotypic characterization of cellular subpopulations. (A-B) Functional enrichment analysis of differentially expressed genes within cell subpopulation. (C) Analysis of the metabolic landscape of cell subpopulations. (D) Cellular polymeric communication network. (E) The contribution of signal strength in both incoming and outgoing signals shapes the landscape of cellular subgroups. (F-G) Analysis of signal transmission into and out of cellular subgroups.

Characteristics of Different Subtypes of Malignant Cells at the Single-Cell Level

To delve deeper into the heterogeneity of the two tumor subtypes, we initially utilized the copykat algorithm to pinpoint potential malignant cell subclusters, including hepatocytes, fibroblasts, and pericellular cells, by detecting whole-genome aneuploidy. This approach facilitated the identification of potential malignant cells, which were subsequently aligned with the prior cell annotation results (Fig.8A). Subsequently, we employed the previously derived multi-omics consensus clustering features to classify the malignant cells into two subtypes: CS1 and CS2. This analysis enabled the identification of differentially expressed genes between the two subtypes. Enrichment analysis of these genes revealed significant differences in various metabolic and signaling pathways between CS1 and CS2 malignant cells, including purine metabolism, pyrimidine metabolism, glycolysis, the PI3K-AKT signaling pathway, and the P53 signaling pathway (Fig.8B). A targeted analysis of the metabolic variances between the CS1 and CS2 subtypes revealed notable differences, indicating significantly higher metabolic activity in CS1 compared to CS2 (Fig.8C). Then, upon integrating the malignant cells classified as CS1 and CS2 into the comprehensive cell communication network, we observed that CS1 and CS2 exhibited widespread regulation of all other cell groups in the TME (Fig.8D). Subsequently, we analyzed the signal output and input patterns of various cellular subgroups and found that the MIF signaling pathway exhibited the highest frequency and greatest contribution in the signal communication network, whether at the input or output end(Fig.8E-G). We further analyzed the output levels of the MIF signaling pathway in various cellular subgroups, focusing on which cells were outputting the MIF signal. The results showed that the CS1 and CS2 subtypes mainly output MIF, with CS1 subtype being relatively predominant(Fig.8H). Further investigation into the regulatory effects of CS1 malignant cells on other cell groups demonstrated that the CS1 subtype exerted extensive regulatory effects on nearly all immune cells in the TME, including CD8+ T cells, macrophages, B cells, and neutrophils, primarily through the MIF- (CD74+CXCR4) pathway (Fig.8I).

Exploring the characteristics of two subtypes of malignant cells at the scRNA-seq level. (A) Mapping of malignant cells in the cell clustering results. (B) Enrichment analysis of differential genes between different subtypes of malignant cells. (C) Metabolic level analysis of different subtypes of malignant cells. (D) Annotation of the cellular subpopulation signaling landscape after malignant cell annotation. (E-G) Analysis of signal input-output patterns in cellular subpopulations. (H) Differential levels of the MIF signaling pathway in cellular subpopulation outputs. (I) Signaling modulation pattern of CS1 subtype malignant cells on other cells.

Discussion

The role of genes in disease development, particularly in tumorigenesis, is well-established. Studying gene expression processes significantly enhances our understanding of tumors. Gene expression is intricately regulated by genetic mechanisms such as methylation and mutations. Therefore, integrating multi-omics data analysis is crucial for gaining deeper insights into disease-specific regulatory mechanisms. However, most studies to date have primarily focused on single omics, such as mRNA and miRNA expression research. Moreover, the choice of omics clustering methods in research often relies on personal preference, exacerbating the limitations of individual approaches in studies. Our research aims to bridge this gap by integrating ten state-of-the-art clustering algorithms to identify two prognostic subtypes with distinct characteristics, which could potentially facilitate precise stratification and treatment of HCC patients. Furthermore, our novel classification method has demonstrated stability across multiple cohorts.

Significant prognostic differences among patients with the same tumor type may be influenced by various factors, including tumor tissue heterogeneity, metabolic reprogramming, signaling patterns, and disparities in the TME. As such, our study undertook initial investigations to elucidate the underpinnings of these prognostic discrepancies across patient subtypes. Initially, we evaluated the mutation landscape across subtypes and observed no notable differences in TMB and FGA between the two subtypes, implying that subtype heterogeneity may not originate from gene mutations. Subsequent analysis of differentially expressed genes and gene enrichment indicated significant enrichment in tumor-related pathways, particularly metabolism and signal transduction pathways like lipid peroxidation, fatty acid metabolism, amino acid metabolism, PI3K-NOX, and EGF-EGFR signaling pathways. This suggests potential variances in metabolic reprogramming and signaling patterns between the subtypes, which can influence patient prognosis by governing the TME. Consequently, we conducted a further analysis of immune cell infiltration in the TME, revealing lower immune cell infiltration in CS2 compared to CS1, implying a potentially lower malignancy in CS2 with reduced release of tumor-associated antigens and subsequent lower immune response and infiltration. Additionally, we thoroughly examined the signaling pathways and biological characteristics closely associated with tumor development. The findings revealed that the CS1 subtype exhibited greater activity than the CS2 subtype in multiple pathways linked to tumorigenesis, including the MAPK, JAK-STAT, NFκB, PI3K-AKT, VEGF, angiogenesis, and hypoxia pathways. These pathways are widely recognized for their close association with tumor development and their significant roles in tumor transformation and progression. Furthermore, the hypoxia pathway and angiogenesis pathway are well-documented to contribute to the formation of an immunosuppressive TME in HCC[19]. Hypoxia directly induces VEGF production[20–21], which can exert inhibitory effects on various immune cells, such as CD8+ T cells[22]. Additionally, VEGF-induced angiogenesis within the TME exacerbates hypoxia, leading to a detrimental feedback loop[23]. These factors likely contribute to the observed heterogeneity and prognostic differences between the two subtypes. Finally, we conducted an in-depth assessment of the impact of subtyping on clinical treatment and prognosis assessment for patients. Our findings revealed that CS2 exhibited enhanced sensitivity to the conventional drug 5-fluorouracil utilized in hepatic arterial infusion chemotherapy, as well as to several tyrosine kinase inhibitors, including gefitinib, afatinib, and erlotinib, compared to CS1. This discovery holds the potential to inform tailored treatment strategies for patients based on their subtype, consequently improving treatment efficacy and prognosis.

The above findings suggest that our classification method reliably distinguishes patients into two subtypes characterized by significant prognostic differences. These prognostic disparities may arise from heterogeneity linked to variations in metabolic reprogramming and signaling patterns, thus providing a theoretical foundation for the classification’s validity. Consequently, we developed a prognostic prediction model based on subtype characteristics to enhance the clinical applicability of the classification and effectively predict patient prognosis. Machine learning algorithms serve as powerful tools for analyzing multi-omics data and constructing prognostic models. We utilized the TCGA-LIHC cohort as the training set and the ICGC-JP cohort as the validation set. After testing 101 algorithm combinations to identify the optimal model, we employed the C-index to evaluate predictive performance, ultimately determining the model with the highest C-index, which signifies superior predictive capability.

The data revealed that the prognostic model, developed using the Random Survival Forest (RSF) algorithm and incorporating 5 hub genes, demonstrated superior performance. Upon assigning risk scores and stratifying patients in both the training and validation sets based on this model, it became evident that individuals in the high-risk category exhibited significantly poorer prognoses compared to those in the low-risk category, affirming the success of our model construction. Subsequently, we compared our model with several previously established prognostic prediction models to predict patient outcomes and assessed their performance in both cohorts. Through comprehensive evaluations, our model consistently outperformed the other models across both sets. Additionally, we conducted further assessments using calibration curves and decision curve analysis, yielding consistent results indicating the robust predictive capabilities of our model for patient prognosis.

After exploring the clinical implications of our classification method, we conducted an analysis of tumor subtype heterogeneity at the single-cell level to uncover potential mechanisms contributing to differences in prognosis and heterogeneity between subtypes. Initially, we annotated a single-cell sequencing dataset comprising tumor tissue samples from 31 HCC patients, effectively categorizing cells into 12 distinct groups. Subsequently, leveraging the CopyKAT algorithm, we analyzed gene expression levels to identify non-diploid cells as malignant. Employing the same classification method used previously, we successfully categorized malignant cells into two subtypes and assessed differential gene expression and metabolic pathways between them. Our findings aligned with RNA-bulk sequencing data, highlighting significant distinctions between subtypes across various tumor-related signaling pathways and metabolic processes.

Our findings indicate significant heterogeneity among tumor cells of the two subtypes. Additionally, we conducted an analysis of signaling patterns between tumor cells and other cell groups. The results revealed that CS1 malignant cells exert substantial regulatory effects on immune cells within the TME, including CD8+ T cells, macrophages, B cells, and neutrophils, mediated by the MIF-(CD74+CXCR4) pathway. Prior research has elucidated that MIF binds to its receptor CD74, leading to its phosphorylation and the recruitment of CD44, triggering downstream phosphorylation signals via the proto-oncogene SRC[24–25]. This cascade encompasses the transcriptional activation of cyclin D1, phosphorylation of retinoblastoma protein, and the involvement of multiple pro-proliferation and anti-apoptotic pathways such as the MAPK/ERK and PI3K/AKT pathways[26–27].

The MIF signaling pathway is intricately involved in diverse cellular processes. It stabilizes the p53-MDM2 complex, promoting cell proliferation and inhibiting apoptosis[28]. There is a synergistic relationship between MIF and HIF-1α, where HIF-1α binds to the 5’-UTR hypoxia response element of MIF, thereby enhancing MIF expression regulated by CREB[29]. Additionally, MIF activates a p53-dependent feedback loop of HIF-1α, crucial for tumor hypoxia adaptation[30]. Moreover, MIF prevents proteasome-mediated degradation of HIF-1α and forms a ternary complex with CSN5 and HIF-1α, amplifying the tumor hypoxia response[31]. Our prior findings indicate heightened activity in MAPK, PI3K-AKT signaling, and the hypoxia pathway in the CS1 subtype compared to CS2. Integrating RNA-bulk and single-cell sequencing data suggests a pivotal role for the MIF pathway in subtype heterogeneity.

In comparison to prior studies, our research exhibits several distinctive characteristics. First, we acknowledged the potential substantial heterogeneity in HCC patients, shaping our research perspective towards offering more precise stratification and treatment for patients. Second, we incorporated 5 types of omics data and employed 10 clustering methods for analysis, ultimately amalgamating them to derive a consensus clustering method, thereby mitigating the impact of clustering method selection bias on our analysis. Third, our modeling was underpinned by the features of consensus clustering subtypes, utilizing 10 machine learning algorithms combined into 101 algorithmic configurations for modeling. We identified the model with the best average C-index performance, surpassing several previously reported prediction models in terms of predictive efficacy. Fourth, subsequent to identifying potential heterogeneity between subtypes in RNA-bulk data analysis, we further delved into single-cell data. Our exploration revealed the involvement of the MIF signaling pathway in tumor heterogeneity and the resulting prognostic disparities in HCC patients. This comprehensive approach allowed us to conduct an extensive study around the consensus clustering method, elucidating both its clinical application value and the mechanisms underpinning the observed heterogeneity and prognostic differences between subtypes.

Conclusion

In summary, our study is based on a new multi-omics clustering consensus typing method. While discovering the prognostic value of this typing method through the analysis of RNA-bulk data and scRNA-seq data, we focused on the tumor heterogeneity caused by metabolic reprogramming and changes in signaling transduction patterns. We initially explored the reasons for the differences in prognosis among patients of different types and found that the MIF signaling pathway is involved in the occurrence and development of HCC. Due to the use of bioinformatics and limited samples and survival data in our study, it is necessary to further validate potential mechanisms through subsequent basic experiments and analysis of larger cohort data.

eLife Assessment

This manuscript offers valuable insights by identifying two distinct liver cancer subtypes through multi-omics integration and developing a robust prognostic model, validated across various datasets, including single-cell RNA sequencing. The evidence is solid, with comprehensive validation in both internal and independent cohorts; however, the reliance on computational methods highlights the necessity for further experimental validation to fully confirm the mechanistic insights.

Reviewer #1 (Public review):

Summary:

The authors aimed to classify hepatocellular carcinoma (HCC) patients into distinct subtypes using a comprehensive multi-omics approach. They employed an innovative consensus clustering method that integrates multiple omics data types, including mRNA, lncRNA, miRNA, DNA methylation, and somatic mutations. The study further sought to validate these subtypes by developing prognostic models using machine learning algorithms and extending the findings through single-cell RNA sequencing (scRNA-seq) to explore the cellular mechanisms driving subtype-specific prognostic differences.

Strengths:

(1) Comprehensive Data Integration: The study's integration of various omics data provides a well-rounded view of the molecular characteristics underlying HCC. This multi-omics approach is a significant strength, as it allows for more accurate and detailed classification of cancer subtypes.

(2) Innovative Methodology: The use of a consensus clustering approach that combines results from 10 different clustering algorithms is a notable methodological advancement. This approach reduces the bias that can result from relying on a single clustering method, enhancing the robustness of the findings.

(3) Machine Learning-Based Prognostic Modeling: The authors rigorously apply a wide array of machine learning algorithms to develop and validate prognostic models, testing 101 different algorithm combinations. This comprehensive approach underscores the study's commitment to identifying the most predictive models, which is a considerable strength.

(4) Validation Across Multiple Cohorts: The external validation of findings in independent cohorts is a critical strength, as it increases the generalizability and reliability of the results. This step is essential for demonstrating the clinical relevance of the proposed subtypes and prognostic models.

Weaknesses:

(1) Inconsistent Storyline:

Despite the extensive data mining and rigorous methodologies, the manuscript suffers from a lack of a coherent and consistent narrative. The transition between different sections, particularly from multi-omics data integration to single-cell validation, feels disjointed. A clearer articulation of how each analysis ties into the overall research question would improve the manuscript.

(2) Questionable Relevance of Immune Cell Activity Analysis:

The evaluation of immune cell activities within the cancer cell model raises concerns about its meaningfulness. The methods used to assess immune function in the tumor microenvironment may not be fully appropriate, potentially limiting the insights gained from this part of the study.

(3) Incomplete Single-Cell RNA-Seq Validation:

The validation of the findings using single-cell RNA-seq data appears insufficient to fully support the study's claims. While the authors make an effort to extend their findings to the single-cell level, the analysis lacks depth. A more comprehensive validation is necessary to substantiate the robustness of the identified subtypes.

(4) Figures and Visualizations:

Several figures in the manuscript are missing necessary information, which affects the clarity of the results. For instance, the pathways in Figure 3A could be clustered to enhance interpretability, the blue bar in Figure 4A is unexplained, and Figure 4B is not discussed in the text. Additionally, the figure legend in Figure 7C lacks detail, and many figure descriptions merely repeat the captions without providing deeper insights.

(5) Appraisal of the Study's Aims and Results:

The authors have set out to achieve an ambitious goal of classifying HCC patients into distinct prognostic subtypes and validating these findings through both bulk and single-cell analyses. While the methodologies employed are innovative and the data integration comprehensive, the study falls short of fully achieving its aims due to inconsistencies in the narrative and incomplete validation. The results partially support the conclusions, but the lack of coherence and depth in certain areas limits the overall impact of the study.

(6) Impact on the Field:

If the identified weaknesses are addressed, this study has the potential to significantly impact the field of HCC research. The multi-omics approach combined with machine learning is a powerful framework that could set a new standard for cancer subtype classification. However, the current state of the manuscript leaves some uncertainty regarding the practical applicability of the findings, particularly in clinical settings.

(6) Additional Context

For readers and researchers, this study offers a valuable look into the potential of integrating multi-omics data with machine learning to improve cancer classification and prognostication. However, readers should be aware of the noted weaknesses, particularly the need for more consistent narrative development and comprehensive validation of the methods. Addressing these issues could greatly enhance the study's utility and relevance to the community.

Reviewer #2 (Public review):

Summary:

Overall, this is a well-executed and insightful study. With some refinement to the presentation and a deeper exploration of the implications, the manuscript will make a significant contribution to the field of cancer genomics and personalized medicine.

Strengths:

The manuscript integrates multi-omics data with machine learning to address the significant heterogeneity of hepatocellular carcinoma (HCC). The use of multiple clustering algorithms and a consensus method strengthens the robustness of the findings. The study successfully develops a prognostic model with excellent predictive accuracy, validated across independent datasets. This adds considerable value to the field, particularly in providing individualized treatment strategies. The identification of two distinct liver cancer subtypes with different biological and metabolic characteristics is well-supported by the data, offering a promising direction for personalized medicine.

Weaknesses:

(1) Consider streamlining the presentation of methods, especially regarding the clustering algorithms and machine learning models. Readers may find it difficult to follow the exact process unless more clearly outlined.

(2) Some figures, such as the signaling pathways and heatmaps, are critical to understanding the study's findings. Ensure that all figures are high quality, easy to interpret, and adequately labeled. You may also want to highlight the key findings within the figure captions more explicitly.

(3) While the manuscript does compare its prognostic model to those previously published, the novelty of the findings could be emphasized more clearly. Discussing the potential limitations of the study (e.g., the reliance on computational models and small sample sizes for scRNA-seq) could strengthen the manuscript.

(4) The manuscript mentions that the data was split into training and validation datasets in a 1:1 ratio. How was the performance verified? Is there an independent test set?

(5) The role of the MIF signaling pathway in subtype differentiation is intriguing, but further mechanistic insights into how this pathway drives the differences between CS1 and CS2 could be discussed in more detail. If experimental evidence for this pathway exists in the literature, it should be mentioned.

(6) Some sentences are quite long and complex, which can affect readability. Breaking them down into shorter, clearer sentences would improve the flow.

Author response:

The following is the authors’ response to the original reviews

Reviewer #1 (Recommendations for the authors):

(1) Storyline and Narrative Flow:

Consider revising the manuscript to create a more coherent and consistent narrative. Clarify how each section of the study-particularly the transition from multi-omics data integration to single-cell RNA-seq validation-contributes to the overall research question. This will help readers better understand the logical flow of the study.

We thank the reviewer’s suggestion, which have highlighted the deficiencies in this area, and we have made appropriate modifications:

We have modified some text, including the connections between different sections in the results part and the objectives and roles of various analyses in each section, thus enhancing the coherence between the contexts and clarifying the objectives and functions of each analysis, We believe this will help readers better understand the main content of the entire text.

(2) Immune Cell Activity Analysis:

Reevaluate the methods used to assess immune cell activities within the context of the tumor microenvironment. Consider providing additional justification for the relevance of using the cancer cell model for this analysis. If necessary, explore alternative methods or models that might offer more meaningful insights into immune-tumor interactions.

We thank the reviewer’s suggestion, which have highlighted the deficiencies in this area, and we have made appropriate modifications:

Using RNA-Bulk data, we evaluated the tumor immune microenvironment through various methods to assess immune infiltration levels and responses to immunotherapy. We found that the results were largely consistent with those presented in the manuscript, providing strong support for our viewpoints. We also acknowledge the limitations of findings from bioinformatics analysis. In our upcoming research, we plan to develop organoid models with gene expression patterns of both CS1 and CS2 subtypes, using these models as a foundation for studying the tumor immune microenvironment.

(3) Single-Cell RNA-Seq Validation:

Expand the validation of your findings using single-cell RNA-seq data. This could include more in-depth analyses that explore the heterogeneity within the subtypes and confirm the robustness of your classification method at the single-cell level. This would strengthen the support for your claims about the relevance of the identified subtypes.

We thank the reviewer’s suggestion, which have highlighted the deficiencies in this area, and we have made appropriate modifications:

In this manuscript, we employed the NTP algorithm to classify malignant cells identified by the CopyKAT algorithm using characteristic genes of CS1 and CS2 subtypes. This approach is similar to previous method that analyzed patients in the ICGC cohort with the same subtype genes. We consider this classification method valid.

After classifying the malignant cells, we performed metabolic and cell communication analyses on the CS1 and CS2 subtype cells, revealing significant differences in biological pathways enriched by differential genes, metabolic levels, and cell signaling patterns. These differences align with variations observed in prior classifications and analyses based on RNA-Bulk data.

We also acknowledge that validating the classification method solely with the single-cell dataset from this study is insufficient. We analyzed GSE202642 using the same processes and methods as GSE229772, finding that the results were generally consistent, indicating that our classification method exhibits a degree of robustness at the single-cell level.

(4) Methodological Justification:

Provide a more detailed rationale for the selection of machine learning algorithms and integration strategies used in the study. Explain why the chosen methods are particularly well-suited for this research, and discuss any potential limitations they might have.

We thank the reviewer’s suggestion, which have highlighted the deficiencies in this area, and we have made appropriate modifications:

We have updated the methodology section to enhance readers' understanding of the fundamental principles involved. This analysis has two key features: first, it combines 10 machine learning algorithms to generate 101 models and ultimately selects the prognostic prediction model with the highest C-index from these 101 algorithms; second, it utilizes the LOOCV method to analyze the training and validation sets. Compared to the conventional method of randomly dividing the training and validation sets by a fixed ratio, this approach significantly minimizes the bias and randomness introduced by the splitting process. Therefore, we believe this analysis can leverage the characteristic genes of the CS1 and CS2 subtypes, combined with existing clinical data from public databases, to yield results that are more accurate and reliable than the commonly used prognostic models in previous literature, such as COX regression and Lasso regression, as well as other individual algorithms. While this analysis presents advantages over some previous modeling methods, it is essential to recognize that it remains based on analyses conducted using public databases, which may obscure certain factors that might be clinically relevant to patient prognosis due to the mathematical logic of the algorithms.

(5) Figures and Visualizations:

Improve the clarity of your figures by addressing the following:

a) Figure 3A: Cluster the pathways to make the comparisons clearer and more meaningful.

b) Figure 4A: Clearly explain the significance of the blue bar.

c) Figure 4B: Ensure this figure is discussed in the main text to justify its inclusion.

d) Figure 7C: Enhance the figure legend to provide more informative details.

Additionally, ensure that figure descriptions go beyond the captions and provide detailed explanations that help the reader understand the significance of each figure.

We thank the reviewer’s suggestion, which have highlighted the deficiencies in this area, and we have made appropriate modifications:

Figure 3A: We clustered the samples based on CS1 and CS2 subtypes and displayed the immune-related cell scores of each sample as a heatmap.

Figure 4A: The blue bars in the figure represent the average C-index of this algorithm combination in the training dataset TCGA and the validation dataset ICGC, which we have supplemented in the corresponding sections of the text.

Figure 4B: We described this figure in the results section, which primarily aims to validate whether our prognostic prediction model can predict patient outcomes in the TCGA cohort. The results showed that after performing prognostic risk scoring on patients based on the prediction model and categorizing them into high-risk and low-risk groups, the two groups exhibited significant prognostic differences, with the high-risk group showing worse outcomes compared to the low-risk group. This indicates that our prognostic prediction model can effectively distinguish the prognostic risk differences among patients in the TCGA-LIHC cohort. We also discussed these findings in the discussion section.

Figure 7C: We used both point color and size to visualize the levels of metabolic scores, resulting in two dimensions in the legend, which actually represent the same information. Therefore, we removed the results that used point size to indicate the levels of metabolic scores.

(6) Supplementary Materials:

Consider including more detailed supplementary materials that provide additional validation data, extended methodological descriptions, and any other information that would support the robustness of your findings.

We thank the reviewer’s suggestion, which have highlighted the deficiencies in this area, and we have made appropriate modifications:

In the subsequent version of the record, we will upload the important results obtained during the research to GitHub, and in this revision, we have updated some figures that may better explain the results or the robustness of the findings as supplementary materials.

(7) Recent Literature:

a) Incorporate more recent studies in your discussion, especially those related to HCC subtypes and the application of machine learning in oncology. This will provide a more current context for your work and help position your findings within the broader field.

We thank the reviewer’s suggestion, which have highlighted the deficiencies in this area, and we have made appropriate modifications:

We have reviewed several studies related to HCC subtype classification and the application of machine learning in this field. In the discussion section, we summarize the significance and limitations of these studies. Additionally, we discuss the characteristics of our study in comparison to previous research in this field.

(8) Data and Code Availability:

Ensure that all data, code, and materials used in your study are made available in line with eLife's policies. Provide clear links to repositories where readers can access the data and code used in your analyses.

We thank the reviewer’s suggestion, which have highlighted the deficiencies in this area, and we have made appropriate modifications:

We have examined the relevant data, code, and materials. We confirm that we have indicated the sources of the data and tools used in the analysis within the manuscript. Moreover, these data and tools are accessible via the websites or references we have provided.

Reviewer #2 (Recommendations for the authors):

(1) While the computational findings are robust, further experimental validation of the two subtypes, particularly the role of the MIF signaling pathway, would strengthen the biological relevance of the findings. In vitro or in vivo validation could confirm the proposed mechanisms and their influence on patient prognosis.

We thank the reviewer’s suggestion, which have highlighted the deficiencies in this area, and we have made appropriate modifications:

We intend to verify our findings in future studies using tumor cell line models and animal models. We aim to identify and intervene with key molecules in the MIF signaling pathway. We will investigate how the MIF signaling pathway affects tumor sensitivity to treatment in both cell line and animal models, along with the underlying mechanisms.

(2) Consider testing the model on additional independent cohorts beyond the TCGA and ICGC datasets to further demonstrate its generalizability and applicability across different patient populations.

We thank the reviewer’s suggestion, which have highlighted the deficiencies in this area, and we have made appropriate modifications:

We analyzed the GSE14520 study recorded in the GEO database, which uploaded a cohort consisting of 209 HCC patients and their corresponding RNA sequencing data. We validated the prognostic model obtained in this study using this cohort, and found that the model effectively distinguishes patients into high-risk and low-risk prognostic categories. Furthermore, there is a significant prognostic difference between the high-risk and low-risk patient groups. This is consistent with the results we obtained previously.

(3) Review the manuscript for long or complex sentences, which can be broken down into shorter, more readable parts.

We have made revisions to the long and complex sentences in the manuscript without compromising its academic integrity and rationality, with the hope that this will help readers better understand the content of this study.

During the revision process, in addition to addressing the reviewer comments, we conducted a thorough review of the analysis. In the course of this review, we identified a few errors in the data usage and have since corrected the relevant data and figures:

Figure 4: Due to space constraints, we adjusted the composition of the figures after incorporating the validation results from the GSE14520 dataset.

Figure 5A: We rechecked the regression coefficients included in the model, updated several more recent prognostic models, and calculated the C-index for 20 prognostic models in the TCGA and ICGC cohorts using a method consistent with previous studies.

Figure 5C-D: We adjusted the clarity of the figures.

Figure 8: We reclassified the selected malignant cells and updated the subtypes results. Subsequently, based on the repeatedly confirmed typing results, we comprehensively updated the analysis results of the subsequent cell communication network construction, ensuring that the entire analysis process remains consistent with previous findings. We also adjusted the composition of the figure and presented the images that could not be conveniently merged due to space constraints as Figure 9.

Version 2

Abstract

Abstract

Purpose

Hepatocellular carcinoma (HCC) is a highly aggressive tumor characterized by significant heterogeneity and invasiveness, leading to a lack of precise individualized treatment strategies and poor patient outcomes. This necessitates the urgent development of accurate patient stratification methods and targeted therapies based on distinct tumor characteristics.

Experimental Design

By integrating gene expression data from The Cancer Genome Atlas (TCGA), International Cancer Genome Consortium (ICGC), and Gene Expression Omnibus (GEO), we identified subtypes through a multi-omics consensus clustering approach amalgamated from 10 clustering techniques. Subsequently, we developed a prognostic model, employing machine learning algorithms, based on subtype classification features. Finally, by analyzing single cell sequencing data, we investigated the mechanisms driving prognostic variations among distinct subtypes.

Results

First, we developed a novel consensus clustering method that categorizes liver cancer patients into two subtypes, CS1 and CS2. Second, we constructed a prognostic prediction model, which demonstrated superior predictive accuracy compared to several models published in the past five years. Finally, we observed differences between CS1 and CS2 in various metabolic pathways, biological processes, and signaling pathways, such as fatty acid metabolism, hypoxia levels, PI3K-AKT and MIF signaling pathway.

Introduction

Liver cancer is currently one of the most common tumors, and among primary liver cancers, hepatocellular carcinoma (HCC) is the most common pathological type. In 2022, liver cancer saw widespread new cases globally, and its incidence rate has been gradually increasing in recent years, ranking 6th among all tumors worldwide[1]. Meanwhile, liver cancer ranks 3rd among tumors causing patient deaths, with a relative 5-year survival rate of about 18%, highlighting the significantly poor prognosis of this disease[2].

With the advancement of current treatment technologies, there are ongoing attempts to employ advanced treatment methods for innovative liver cancer therapy. Nevertheless, up to now, the progress of strategies aimed at treating liver cancer, especially unresectable liver cancer, has been limited. For instance, targeted therapies and immunotherapies have been the focus of extensive research efforts and have shown promising results in some patients. However, a significant number of patients still do not obtain clinical benefits from these approaches[3]. This may be due to the significant heterogeneity of liver cancer; classifying liver cancer patients according to molecular subtypes might help overcome this difficulty[4]. This urgently necessitates leveraging large-scale multi-omics data and advanced machine learning algorithms to identify biomarkers capable of efficiently predicting clinical treatment outcomes for liver cancer patients.

In this study, we combined mRNA, long non-coding RNA (lncRNA), and microRNA (miRNA) expression profiles, genomic mutations, and epigenomic DNA methylation data, using a consensus strategy integrating 10 multi-omics clustering methods to develop comprehensive consensus subtypes of liver cancer, and conducted preliminary exploration of the molecular characteristics and potential clinical applications of the two subtypes. Subsequently, we further expanded the clinical application value of the subtype. Based on the gene expression profiles of two subtypes, we used a combination of 101 machine learning algorithms to construct the best-performing prognostic model. This prognostic model showed excellent prognostic value in both training, internal and external validation sets, and outperformed several other prognostic models in comparisons. Finally, we validated the subtypes at the single-cell sequencing cell level using the same typing method, and further explored the possible causes and mechanisms of prognostic differences in patients with different subtypes.

Materials and methods

Data collection and processing

Download the molecular profiles of the TCGA-LIHC cohort dataset from the TCGA database, including 339 LIHC cases with complete transcriptome expression, somatic mutations, copy number alterations (CNA), DNA methylation, and clinical information. The transcriptome profiles of mRNA and lncRNA were obtained using the “TCGAbiolinks” R package[5], and GENCODE27 was used to convert Ensemble IDs to gene symbols; Somatic mutation data were also obtained using “TCGAbiolinks” R package and processed with the “maftools” R package[6]; DNA methylation profiles and clinical information were downloaded from UCSC Xena (https://xenabrowser.net/); Meanwhile, we also downloaded 445 LIHC cases with complete transcriptome expression and clinical information from the LIHC.JP (https://dcc.icgc.org/) and GSE14520 (https://www.ncbi.nlm.nih.gov/geo/, accession number GSE14520) cohorts; The single-cell sequencing datasets are sourced from the GEO database (https://www.ncbi.nlm.nih.gov/geo/, accession number GSE229772).

Construct multi-omics consensus clustering classification and perform validation

In this study, we utilized the “getElites” function from the multi-omics integration and visualization (MOVICS) package to screen gene features in cancer subtypes[7]. For continuous variables (mRNA, lncRNA, miRNA, and methylation), we set the “method” parameter of the “getElites” function to “mad” to screen the top 1000 genes with the greatest variability; For binary variable gene mutation data, we conducted screening based on mutation frequency by setting the “method” parameter to “freq” to identify genes with a mutation frequency exceeding 10. The results from these five dimensions were incorporated into our study for further analysis. After the initial feature selection, we used the “getClustNum” function in the MOVICS package to further determine the optimal number of clusters for our study and ultimately decided to divide them into two subtypes. Next, we applied the “getMOIC” function for clustering analysis. We used 10 clustering algorithms (CIMLR, ConsensusClustering, SNF, iClusterBayes, PINSPlus, moCluster, NEMO, IntNMF, COCA, and LRA) as input for the “methods.list” parameter and employed the default parameters provided by the MOVICS package. As a result, we obtained the clustering results for each method. After calculating the clustering results of the 10 methods, we integrated the results of different algorithms based on the concept of consensus clustering using the “getConsensusMOIC” function, obtaining the final clustering results. Subsequently, we conducted nearest template prediction(NTP) analysis, performed subtyping in the external ICGC-JP cohort, and validated the subtyping method.

Molecular Characteristics of Multi-Omics Consensus Clustering Subtypes

We analyzed the mutation status using the “maftools” R package, then used the limma method to analyze the differentially expressed genes between the two subtypes[8], and performed enrichment analysis with the clusterProfiler R package[9].

Cell, signaling pathways, stemness, and other phenotypic characteristics of multi-omics consensus clustering subtypes

Using ssGSEA analysis, we calculated the enrichment scores of different immune cell characteristics between the two subtypes based on immune cell gene sets reported in the literature[10]; Next, we analyzed the activation status of cancer pathways and some biological processes in patients using the GSEA algorithm, and analyzed the sensitivity of the two subtypes to conventional therapeutic drugs. Both of these steps were performed using the “GSVA” R package[11]. The expression data from CCLE were sourced from the Broad Institute Cancer Cell Line Encyclopedia (https://sites.broadinstitute.org/ccle/); The CTRP v.2.0 (https://portals.broadinstitute.org/ctrp) and PRISM dataset (24Q2; https://depmap.org/portal/prism/) were utilized to acquire drug sensitivity data; The half-maximal inhibitory concentration (IC50) was used as an indicator of drug sensitivity.

Constructing a prognostic prediction model based on the consensus clustering of multi-omics features

To build the prognostic model, we integrated 10 machine learning algorithms and 101 algorithm combinations[12]. The integrative algorithms included random survival forest (RSF), elastic network (Enet), Lasso, Ridge, stepwise Cox, CoxBoost, partial least squares regression for Cox (plsRcox), supervised principal components (SuperPC), generalised boosted regression modelling (GBM), and survival support vector machine (survival-SVM). The signature generation procedure was as follows: (a) Univariate Cox regression identified prognostic genes in the TCGA-LIHC; (b) Then, 101 algorithm combinations were performed on the prognostic genes to fit prediction models based on the leave-one-out cross-validation (LOOCV) framework in the TCGA-LIHC cohort and incorporated 10 machine learning algorithms, including CoxBoost, stepwise Cox, Lasso, Ridge, Enet, survival-SVMs, GBMs, SuperPC, plsRcox, and RSF, creating a total of 101 different combinations of these algorithms. Ultimately, these 101 algorithm combinations will be utilized to develop the most predictive prognostic model with the best C-index performance. We then scored each sample in the training and validation sets based on the model and classified the samples into high-risk and low-risk groups according to the optimal cut-off value of scores, and the “survminer” R package (https://github.com/kassambara/survminer/tree/master) was used to determine the optimal cut-off value. Cox regression and Kaplan–Meier analyses were performed via the “survival” R package (https://CRAN.R-project.org/package=survival).

Comparison of Prognostic Prediction Models

We evaluated the prognostic predictive ability of all models included in the study using the C-index in each cohort. To enhance the clinical utility of the model, we constructed a nomogram based on factors obtained from multivariable Cox regression. We plotted time-dependent C-index curves and calibration curves to describe accuracy and used decision curve analysis to calculate the clinical benefit for patients.

scRNA-seq Data Processing

We utilized the “Seurat” R package to process single-cell sequencing data from the GSE229772 dataset[13]. We removed samples with over 10% mitochondrial genes and over 5% ribosomal genes, and filtered out genes expressed in fewer than 3 cells. Next, we identified the most variable genes using the FindVariableFeatures function from the Seurat R package, conducted dimensionality reduction via principal component analysis (PCA), and identified a total of 13 clusters through t-SNE clustering analysis. Finally, we annotated the cell populations by combining the expression of characteristic genes in each cell cluster with the expression of commonly used cell-type-specific genes across the clusters.

Characteristic Analysis of Different Cell Subpopulations

We used the “irGSEA” R package[14], employing 6 methods (AUCell, UCell, singscore, ssGSEA, JASMINE, and viper) for pathway enrichment analysis among different cell populations, and performed a comprehensive evaluation of the differential analysis results using robust rank aggregation (RRA) to identify gene sets significantly enriched in most enrichment analysis methods. Moreover, we additionally utilized the “scMetabolism” R package to compare the metabolic difference across various cell types[15]. The metabolic activity of each cell was quantified through integrating the gene sets contained in the KEGG database. Subsequently, we used the “CellChat” R package to explore the interaction patterns among cell populations[16], defining ligands and receptors as outgoing and incoming signals, respectively.

Characteristics of Different Subtypes of Malignant Cells at the Single-Cell Level

We applied the “CopyKAT” R package to analyze gene expression levels across all cells[17], identifying diploid cells as normal cells, with normal and tumor cells separated due to differences in gene expression distributions. Subsequently, we classified tumor cells based on the previously obtained multi-omics consensus clustering approach, and used the “irGSEA” R package to perform pathway enrichment analysis on different subtypes, conducted metabolic differential analysis with the “scMetabolism” R package, and ultimately explored the interaction patterns of different subtype malignant cells with other cell populations using the “CellChat” package. In addition, we analyzed the scRNA-seq dataset GSE202642 using the same workflow as GSE229772, thereby validating important outcomes such as the robustness of the multi-omics clustering consensus typing method at the scRNA data level, the metabolic differences between different subtypes of malignant cells, and the variations in cell communication patterns.

Results

Construct multi-omics consensus clustering classification and perform validation

We utilized clustering prediction indicators, gap statistical analysis, silhouette scores, and insights from previous research involving 10 multi-omics integrated clustering algorithms to ascertain the number of subtypes (Fig.1A). Using a consensus integration method, we linked the clustering results to distinct molecular expression patterns of the transcriptome (mRNA, lncRNA, and miRNA), epigenetic methylation, and somatic mutations, thereby identifying two subtypes (CS1 and CS2) independently (Fig.1B). Significant differences in mRNA, lncRNA, miRNA, DNA CpG methylation sites, and mutation gene characteristics were observed between the two subtypes (Fig.1C). The five types of omics data used in the construction of the multi-omics consensus clustering classification can be found in the supplementary materials. This classification method is strongly correlated with overall survival (p < 0.001), with patients of the CS2 subtype exhibiting better prognostic survival compared to those of the CS1 subtype (Fig.1D). These characteristics suggest that our classification method can effectively distinguish between patients. Subsequently, employing the NTP method, we selected 200 genes specifically upregulated in each subtype as classifiers to validate the classification in an independent external validation set, ICGC, and found a high degree of consistency in the classification results. These 200 feature genes, along with the gene expression profiles and clinical data from the ICGC cohort that we used, can be found in the supplementary materials. Patients with CS1 and CS2 subtypes in the ICGC cohort exhibited similar prognostic differences (Fig.1E-G). This demonstrates the versatility and practical value of our classification method. We will explore the reasons for the significant prognostic differences between the two subtypes of patients by integrating multi-omics data. It will primarily focus on the sensitivity of tumors to current mainstream treatment options, including immunotherapy, targeted therapy, and chemotherapy.

Multi-omics consensus clustering classification.

(A) Calculation of optimal number of subtypes based on cluster prediction index and gap-statistics. (B) Heatmap of the consistency of the integration of the results of the 10 typing algorithms. (C-D) Significant multiomics characterization and prognostic differences between subtypes. (E-G) Validation of consistency of typing results and prognostic value in an external cohort.

Molecular Characteristics of Multi-Omics Consensus Clustering Subtypes

Based on the correlation between tumor mutation burden (TMB) and immune therapy response, we analyzed the gene mutation landscape of two subtypes and found that both subtypes exhibited a high frequency of mutations in TP53, KMT2D, and DNAH6 genes(Fig.2A). Then we analyzed the TMB of the two subtypes of patients to assess the impact of immune therapy responsiveness on the prognosis of these subtypes. The results showed no significant difference in TMB between the two patient subtypes. (Fig.2B-C). Subsequently, we analyzed the bulk RNA sequencing data of patients within the two subtypes, identified differentially expressed genes, and conducted enrichment analysis on these genes. The results revealed significant differential enrichment in several biological pathways, including those related to metabolism and signal transduction, such as lipid peroxidation, fatty acid metabolism, and amino acid metabolism, suggesting notable biological differences between the tumors and associated tissues of the two subtypes (Fig.2D-E). The list of differentially expressed genes can be found in the supplementary materials. Previous studies have reported that heterogeneity among different tumor tissues of the same tumor type largely stems from metabolic reprogramming and changes in signal transduction pathway activity[18–19]. This implies that the heterogeneity between the tumor tissues of the two subtypes may also arise from these factors. Finally, we compared the consistency of subtyping methods with the pathologic staging of patients(Fig.2F).

Differences in molecular characterizaiion of subtypes.

(A) Differences in mutation landscapes of subtypes. (B-C) Genomic characterization from TMB and FGA of subtypes. (D-E) Functional emichment of up and down-regulated differential genes between subtypes. (F) Concordance between subtypes and patients’ pathologic staging.

Cell, signaling pathways, stemness, and other phenotypic characteristics of multi-omics consensus clustering subtypes

In the results presented in Fig. 2, we found that there may be no significant difference in tumor mutation burden between the two subtypes. Therefore, we analyzed the tumor immune microenvironment factors that have been frequently reported to potentially affect the response of tumors to immunotherapy. We applied the ssGSEA algorithm to assess the enrichment status of immune-related cells in the tumor microenvironment (TME) of patients from the TCGA cohort, based on a gene set of common immune-related cell characteristics compiled from multiple literature sources. (Fig.3A). The gene set utilized for this analysis was compiled and summarized from various literature sources. After the preliminary analysis of the immune response of the two tumor subtypes to immunotherapy, we utilized the drug intervention data from the pRRophetic database to assess the sensitivity of the two tumor subtypes to common liver cancer therapies, including chemotherapy and targeted therapy. We found the CS2 subtype exhibited higher sensitivity to 5-fluorouracil, Gefitinib, Afatinib, and Erlotinib compared to the CS1 subtype (Fig.3B). Among these, 5-fluorouracil is the main drug used in hepatic arterial infusion chemotherapy for hepatocellular carcinoma (HCC) patients, whereas Gefitinib, Afatinib, and Erlotinib are tyrosine kinase inhibitors (TKIs) that primarily inhibit epidermal growth factor receptor (EGFR). We further compared the level of infiltration of each cell taxon in subtypes in the TME(Fig.3C). Within the TME of the CS2 subtype, associated with a comparatively more favorable prognosis, nearly all assessed cell distribution levels demonstrated a decrease in comparison to those in the CS1 subtype. This encompasses traditionally recognized inhibitory immune cells like MDSCs and Regulatory T cells, alongside pivotal tumor-killing immune cells like Activated CD8+ T cells and NK cells. In addition, we also conducted similar immune infiltration assessments using ESTIMATE and EPIC scores. The results indicated that the findings from these methods were consistent with those obtained using ssGSEA (Fig. S1A-B). These findings may imply reduced levels of tumor-associated antigens within CS2 tumor tissue, leading to a comparatively subdued activation and induction of immune cells within the TME. After discovering significant differences in the responses of the two tumor subtypes to mainstream liver cancer treatment methods, we employed the ssGSEA method, combined with gene sets corresponding to pathways, to preliminarily assess several common pathways that may influence the tumor’s response to treatment. We discovered that the CS1 subtype is more active than the CS2 subtype in multiple pathways associated with tumor development and progression, including the MAPK, JAK-STAT, NFκB, PI3K-AKT, and VEGF signaling pathways, as well as pathways involved in angiogenesis and hypoxia. No significant differences were observed in the cAMP signaling pathway or the mRNAsi stemness score (Fig. 3D).

Phenotypic featmes of subtypes that may be associated with tumor development.

(A) Landscape of subtype Th1E immune infiltration. (B) Differences in subtype responsiveness to different drugs. (C) Comparison of immune cell infiltration levels in subtypes ofTME. (D) Comparison ofhepatocellular carcinoma-related signaling pathways and biological process activity in subtypes.

Constructing a prognostic prediction model based on the consensus clustering of multi-omics features

After obtaining a classification method with clinical prognostic predictive value and gaining some theoretical support from RNA bulk data analysis, we constructed a patient prognosis prediction model based on this classification method to achieve a more accurate and reasonable prediction of patient outcomes. We selected the top 30 most significantly differentially expressed genes from CS1/CS2 gene templates as candidate prognostic risk genes. Subsequently, we applied 101 algorithmic approaches to determine the correlation and impact weights between these 30 genes and prognosis, leading to the construction of a prognostic prediction model. Simultaneously, we calculated the average C index of each model to assess the predictive capability across all models. The numbers represented by the blue bars indicate the average C-index of the model in both the training and internal validation sets (Fig. 4A). Subsequently, we obtained the risk coefficients of the five genes used to build the RSF model by stepwise Cox regression analysis (Fig.4B). We then employing this model to evaluate patients’ prognostic risk within TCGA-LIHC cohort and stratifying them based on the optimal cutoff value, we observed a notably poorer prognosis among high-risk group patients compared to their low-risk counterparts. The 5 central genes, and the Hazard Ratio associated with these genes in the TCGA-LIHC cohort is depicted in Fig. 4C. Additionally, we illustrated the patients’ prognostic outcomes and variations in central genes as the prognostic risk escalated (Fig. 4D). Subsequently, we validated the model using independent external validation set, the ICGC and GSE14520 cohorts. A notable disparity in prognosis between the two groups was evident, aligning with the outcomes observed in the TCGA-LIHC cohort (Fig. 4E-F). In addition, we also plotted 1-, 2-, 3-year AUC curves in the TCGA-LIHC, ICGC and GSE14520 cohorts (Fig. S1C-E).

Building prognostic prediction models through machine learning.

(A) 101 machine learning algorithms to build screening optimal prognostic models(TCGA1=training set,TCGA2=internal validation set). (B) Regression coefficients of 5 genes obtained in stepwise Cox regression.

Comparison of Prognostic Prediction Models

With the rapid advancement and widespread adoption of Next Generation Sequencing (NGS) technology, numerous prognostic prediction models based on gene expression have been extensively reported. After developing our model, we reviewed and collected prognostic prediction models reported in the past five years and included 19 of these models, which are based on various foundations such as immune infiltration, glucose metabolism, and lipid metabolism. We calculated the concordance index for a total of 20 models in both TCGA-LIHC dataset and ICGC dataset. The average C-index of these models was derived to evaluate the prognostic predictive performance of the model. The results indicated that our model demonstrated superior prognostic prediction performance compared to all other models in both cohorts (Fig. 5A). Subsequently, we incorporated clinical features related to prognosis into the model to construct a nomogram (Fig. 5B). The calibration curve demonstrated that the nomogram had predictive accuracy more consistent with actual conditions (Fig. 5C). Decision curve analysis (DCA) indicated that both the nomogram and the original model could benefit patients, with the nomogram performing slightly better (Fig. 5D).

Comparison and refinement of prognostic prediction models.

(A) Comparison of predictive performance of prognostic models with some prognostic models published in the last 5 years. (B) Constructing a nomogram incorporating patient clinical characteristics. (C-D) Evaluating model performance based on calibration curve and decision curve analysis.

scRNA-seq Data Processing

Given the significant prognostic differences between patients with CS1 and CS2, as well as the notable heterogeneity observed at the cellular and molecular levels in tumor tissues from the RNA-Bulk data analysis, we attempted to further explore the potential mechanisms underlying these differences between the two subtypes by analyzing scRNA-seq data. The GSE229772 dataset from the GEO database was chosen for our analysis. Following quality control filtering, 56,937 cells were derived from 31 samples. By employing hierarchical and unsupervised clustering methods, we delineated cell subclusters according to their expression profiles, categorizing the cells into 13 subclusters (Fig. 6A). Subsequently, we employed established marker genes to profile various cell types and assign annotations, discerning 12 primary cell subgroups: B cells, stromal cells, vascular endothelial cells, CD8+ T cells, epithelial cells, fibroblasts, germ cell-like cells, hepatic stellate cells, hepatocytes, macrophages, neutrophils, and T cells (Fig. 6B-D). Within these, the immune cell group prevailed as the most abundant, succeeded by stromal cells and fibroblasts (Fig. 6E).

ScRNA-seq cell s01iing and annotation.

(A-C) Cellular subpopulation delineation and annotation based on cellular characterization genes. (D-E) Visualization of the proportions of each cell subgroup. (F) Differentially expressed genes by cellular subpopulations.

Characteristic Analysis of Different Cell Subpopulations

To investigate pathway heterogeneity among different cell subclusters, we identified differentially expressed characteristic genes within these subclusters. Subsequently, we employed multiple algorithms, including AUCell, UCell, singscore, ssgsea, JASMINE, and viper, to assess pathway activity variation. Differential analysis results were comprehensively evaluated using robust rank aggregation (RRA), and gene sets significantly enriched across multiple enrichment analysis methods were filtered to obtain consensus results. Statistically significant differentially enriched pathways from these consensus results were visualized (Fig.7A-B). We observed significant differences in several metabolic and signaling pathways among different cell subclusters. Given the reported impact of metabolic changes and signaling pathway variations on the TME and their implications for heterogeneity, we specifically analyzed these pathways. Utilizing the VISION algorithm, we assessed metabolic levels across each cell subcluster and visualized the differences in a dot plot (Fig.7C). Our findings indicated pronounced metabolic variations among cell subclusters, with germ cell-like cells and hepatocytes notably exhibiting heightened metabolic activity compared to other cell groups. Subsequently, we utilized the CellChat R package’s integrated databases for secreted signaling, ECM receptor signaling, cell-cell contact, and non-protein signaling to explore cell interactions within the single-cell dataset. Initially, we examined the cell communication landscape between distinct cell groups and constructed a cell communication network (Fig.7D). Our analysis revealed robust communication links between various cell groups, particularly notable between fibroblasts and hepatic stellate cells, which exhibited signaling connections with nearly all other cells. Then, we visualized the signaling output and input patterns (Fig.7E-G) to illustrate the contributions of different cell groups at both ends.

Phenotypic characterization of cellular subpopulation.

(A-B) Functional enrichrnent analysis of differentially expressed genes within cell subpopulation. (C) Analysis of the metabolic landscape of cell subpopulations. (D) Cellular polymeric communication network. (E) The contribution of signal strength in both incoming and outgoing signals shapes the landscape of cellular subgroups. (F-G) Analysis of signal transmission into and out of cellular subgroups.

Characteristics of Different Subtypes of Malignant Cells at the Single-Cell Level

To delve deeper into the heterogeneity of the two tumor subtypes, we initially utilized the copykat algorithm to pinpoint potential malignant cell subclusters, including hepatocytes, fibroblasts, and hepatic stellate cells, by detecting whole-genome aneuploidy. This approach facilitated the identification of potential malignant cells, which were subsequently aligned with the prior cell annotation results (Fig.8A). Subsequently, we employed the previously derived multi-omics consensus clustering features to classify the malignant cells into two subtypes, CS1 and CS2, by NTP method (Fig.S2). This analysis enabled the identification of differentially expressed genes between the two subtypes. Enrichment analysis of these genes revealed significant differences in various signaling pathways between CS1 and CS2 malignant cells, including the PI3K-AKT and P53 signaling pathway (Fig.8B). A targeted analysis of the metabolic variances between the CS1 and CS2 subtypes revealed notable differences, indicating significantly higher metabolic activity in CS1 compared to CS2 (Fig.8C). Then, upon integrating the malignant cells classified as CS1 and CS2 into the comprehensive cell communication network (Fig.S3), we observed that CS1 and CS2 exhibited widespread regulation of all other cell groups in the TME (Fig.8D). Subsequently, we analyzed the signal output and input patterns of various cellular subgroups and found that the MIF signaling pathway exhibited the highest frequency and greatest contribution in the signal communication network, whether at the input or output end(Fig.8E-G). We further analyzed the output levels of the MIF signaling pathway in various cellular subgroups, focusing on which cells were outputting the MIF signal. The results showed that the CS1 and CS2 subtypes mainly output MIF, with CS1 subtype being relatively predominant(Fig.8H). Further investigation into the regulatory effects of CS1 malignant cells on other cell groups demonstrated that the CS1 subtype exerted extensive regulatory effects on nearly all immune cells in the TME, including CD8+ T cells, macrophages, B cells, and neutrophils, primarily through the MIF- (CD74+CXCR4) pathway (Fig.9). It is noteworthy that during our analysis of GSE202642, we found that the metabolic differences and cell communication pattern variations among different subtypes of malignant cells were largely consistent with the results obtained from the analysis of GSE229772 (Fig.S4-6).

Exploring the characteristics of two subtypes of malignant cells at the scRNA-seq level.

(A) Mapping of malignant cells in the cell clustering results. (B) Enrichment analysis of differential genes between different subtypes of malignant cells. (C) Metabolic level analysis of different subtypes of malignant cells. (D) Annotation of the cellular subpopulation signaling landscape after malignant cell annotation. (E-G) Analysis of signal input-output patterns in cellular subpopulation. (H) Differential levels of the MIF signaling pathway in cellular subpopulation outputs.

Signaling modulation pattern of CSl subtype malignant cells on other cells.

Discussion

The role of genes in disease development, particularly in tumorigenesis, is well-established. Studying gene expression processes significantly enhances our understanding of tumors. Gene expression is intricately regulated by genetic mechanisms such as methylation and mutations. Therefore, integrating multi-omics data analysis is crucial for gaining deeper insights into disease-specific regulatory mechanisms. However, most studies to date have primarily focused on single omics, such as mRNA and miRNA expression research. Moreover, the choice of omics clustering methods in research often relies on personal preference, exacerbating the limitations of individual approaches in studies. Our research aims to bridge this gap by integrating ten state-of-the-art clustering algorithms to identify two prognostic subtypes with distinct characteristics, which could potentially facilitate precise stratification and treatment of HCC patients. Furthermore, our novel classification method has demonstrated stability across multiple cohorts.

Significant prognostic differences among patients with the same tumor type may be influenced by various factors, including tumor tissue heterogeneity, metabolic reprogramming, signaling patterns, and disparities in the TME. As such, our study undertook initial investigations to elucidate the underpinnings of these prognostic discrepancies across patient subtypes. Initially, we evaluated the mutation landscape across subtypes and observed no notable differences in TMB and FGA between the two subtypes, implying that subtype heterogeneity may not originate from gene mutations. Subsequent analysis of differentially expressed genes and gene enrichment indicated significant enrichment in tumor-related pathways, particularly metabolism and signal transduction pathways like lipid peroxidation, fatty acid metabolism, amino acid metabolism, PI3K-NOX, and EGF-EGFR signaling pathways. This suggests potential variances in metabolic reprogramming and signaling patterns between the subtypes, which can influence patient prognosis by governing the TME. Consequently, we conducted a further analysis of immune cell infiltration in the TME, revealing lower immune cell infiltration in CS2 compared to CS1, implying a potentially lower malignancy in CS2 with reduced release of tumor-associated antigens and subsequent lower immune response and infiltration. Additionally, we thoroughly examined the signaling pathways and biological characteristics closely associated with tumor development. The findings revealed that the CS1 subtype exhibited greater activity than the CS2 subtype in multiple pathways linked to tumorigenesis, including the MAPK, JAK-STAT, NFκB, PI3K-AKT, VEGF, angiogenesis, and hypoxia pathways. These pathways are widely recognized for their close association with tumor development and their significant roles in tumor transformation and progression. Furthermore, the hypoxia pathway and angiogenesis pathway are well-documented to contribute to the formation of an immunosuppressive TME in HCC[20]. Hypoxia directly induces VEGF production[21–22], which can exert inhibitory effects on various immune cells, such as CD8+ T cells[23]. Additionally, VEGF-induced angiogenesis within the TME exacerbates hypoxia, leading to a detrimental feedback loop[24]. These factors likely contribute to the observed heterogeneity and prognostic differences between the two subtypes. Finally, we conducted an in-depth assessment of the impact of subtyping on clinical treatment and prognosis assessment for patients. Our findings revealed that CS2 exhibited enhanced sensitivity to the conventional drug 5-fluorouracil utilized in hepatic arterial infusion chemotherapy, as well as to several tyrosine kinase inhibitors, including gefitinib, afatinib, and erlotinib, compared to CS1. This discovery holds the potential to inform tailored treatment strategies for patients based on their subtype, consequently improving treatment efficacy and prognosis.

The above findings suggest that our classification method reliably distinguishes patients into two subtypes characterized by significant prognostic differences. These prognostic disparities may arise from heterogeneity linked to variations in metabolic reprogramming and signaling patterns, thus providing a theoretical foundation for the classification’s validity. Consequently, we developed a prognostic prediction model based on subtype characteristics to enhance the clinical applicability of the classification and effectively predict patient prognosis. Machine learning algorithms serve as powerful tools for analyzing multi-omics data and constructing prognostic models. We utilized the TCGA-LIHC cohort as the training and validation set. After testing 101 algorithm combinations to identify the optimal model, we employed the C-index to evaluate predictive performance, ultimately determining the model with the highest C-index, which signifies superior predictive capability. In addition to selecting the best results from 101 algorithm combinations to achieve greater accuracy and reliability, the use of the LOOCV method can significantly reduce the bias and randomness that arise during the application of the training and validation sets, compared to previous methods that directly split the training and validation sets by a certain ratio. The data revealed that the prognostic model, developed using the Random Survival Forest (RSF) algorithm and incorporating 5 hub genes, demonstrated superior performance. Upon assigning risk scores and stratifying patients in both the training and validation sets based on this model, it became evident that individuals in the high-risk category exhibited significantly poorer prognoses compared to those in the low-risk category, affirming the success of our model construction. Subsequently, we compared our model with 19 previously established prognostic prediction models to predict patient outcomes and assessed their performance in both cohorts. Through comprehensive evaluations, our model consistently outperformed the other models across both sets. Additionally, we conducted further assessments using calibration curves and decision curve analysis, yielding consistent results indicating the robust predictive capabilities of our model for patient prognosis. It should be noted that although the model building method we used has obvious advantages compared with some early modeling methods, it still belongs to the mathematical analysis based on public databases. Therefore, during the analysis process, the mathematical logic of the algorithm inevitably obscures some factors that may be clinically relevant to the prognosis of patients.

After exploring the clinical implications of our classification method, we conducted an analysis of tumor subtype heterogeneity at the single-cell level to uncover potential mechanisms contributing to differences in prognosis and heterogeneity between subtypes. Initially, we annotated a single-cell sequencing dataset comprising tumor tissue samples from 31 HCC patients, effectively categorizing cells into 12 distinct groups. Subsequently, leveraging the CopyKAT algorithm, we analyzed gene expression levels to identify non-diploid cells as malignant. Employing the same classification method used previously, we successfully categorized malignant cells into two subtypes and assessed differential gene expression and metabolic pathways between them. Our findings aligned with RNA-bulk sequencing data, highlighting significant distinctions between subtypes across various tumor-related signaling pathways and metabolic processes.

Our findings indicate significant heterogeneity among tumor cells of the two subtypes. Additionally, we conducted an analysis of signaling patterns between tumor cells and other cell groups. The results revealed that CS1 malignant cells exert substantial regulatory effects on immune cells within the TME, including CD8+ T cells, macrophages, B cells, and neutrophils, mediated by the MIF-(CD74+CXCR4) pathway. Prior research has elucidated that MIF binds to its receptor CD74, leading to its phosphorylation and the recruitment of CD44, triggering downstream phosphorylation signals via the proto-oncogene SRC[25–26]. This cascade encompasses the transcriptional activation of cyclin D1, phosphorylation of retinoblastoma protein, and the involvement of multiple pro-proliferation and anti-apoptotic pathways such as the MAPK/ERK and PI3K/AKT pathways[27–28].

The MIF signaling pathway is intricately involved in diverse cellular processes. It stabilizes the p53-MDM2 complex, promoting cell proliferation and inhibiting apoptosis[29]. There is a synergistic relationship between MIF and HIF-1α, where HIF-1α binds to the 5’-UTR hypoxia response element of MIF, thereby enhancing MIF expression regulated by CREB[30]. Additionally, MIF activates a p53-dependent feedback loop of HIF-1α, crucial for tumor hypoxia adaptation[31]. Moreover, MIF prevents proteasome-mediated degradation of HIF-1α and forms a ternary complex with CSN5 and HIF-1α, amplifying the tumor hypoxia response[32]. Our prior findings indicate heightened activity in MAPK, PI3K-AKT signaling, and the hypoxia pathway in the CS1 subtype compared to CS2. Integrating RNA-bulk and single-cell sequencing data suggests a pivotal role for the MIF pathway in subtype heterogeneity.

We have reviewed and summarized several articles that have been published regarding the classification of HCC patients and the construction of prognostic models using machine learning. A considerable number of studies focus on specific pathways that may affect the tumor microenvironment, metabolic reprogramming, or patient prognosis (such as fatty acid metabolism, ferroptosis, and autophagy), using the expression levels of genes in these pathways as classification features for HCC patients and constructing prognostic models based on general regression analysis. Although this research method can produce classification results related to patient prognosis and generate prognostic models, it also has some issues. First, focusing on a specific pathway from the outset of the study may exclude most classification features outside the designated pathway, leading to limitations in the research. Second, these studies only use the mRNA expression levels of genes to construct classification features, making the utilization of omics data rather singular, which may affect the robustness of the classification method. However, a positive aspect is that some studies have selected key genes associated with prognosis and conducted experimental validation, which is helpful for exploring how classification methods can effectively distinguish patient prognosis.

In comparison to prior studies, our research exhibits several distinctive characteristics. First, we acknowledged the potential substantial heterogeneity in HCC patients, shaping our research perspective towards offering more precise stratification and treatment for patients. Second, we incorporated 5 types of omics data and employed 10 clustering methods for analysis, ultimately amalgamating them to derive a consensus clustering method, thereby mitigating the impact of clustering method selection bias on our analysis. Third, our modeling was underpinned by the features of consensus clustering subtypes, utilizing 10 machine learning algorithms combined into 101 algorithmic configurations for modeling. We identified the model with the best average C-index performance, surpassing several previously reported prediction models in terms of predictive efficacy. Fourth, subsequent to identifying potential heterogeneity between subtypes in RNA-bulk data analysis, we further delved into single-cell data. Fifth, we employed various methods to validate the key results, including the validation of multi-omics profiling features using the ICGC cohort and two scRNA-seq datasets, as well as validating the key results obtained from the scRNA-seq data analysis with GSE202642, which will significantly strengthen the robustness of our results. Our exploration revealed the involvement of the MIF signaling pathway in tumor heterogeneity and the resulting prognostic disparities in HCC patients. This comprehensive approach allowed us to conduct an extensive study around the consensus clustering method, elucidating both its clinical application value and the mechanisms underpinning the observed heterogeneity and prognostic differences between subtypes.

Conclusion

In summary, our study is based on a new multi-omics clustering consensus typing method. While discovering the prognostic value of this typing method through the analysis of RNA-bulk data and scRNA-seq data, we focused on the tumor heterogeneity caused by metabolic reprogramming and changes in signaling transduction patterns. We initially explored the reasons for the differences in prognosis among patients of different types and found that the MIF signaling pathway is involved in the occurrence and development of HCC. Due to the use of bioinformatics and limited samples and survival data in our study, it is necessary to further validate potential mechanisms through subsequent basic experiments and analysis of larger cohort data.